---
title: "01. Data Import and Quality Control"
subtitle: "GSE125969 ependymoma cohort (Gillen et al. 2020)"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

# Context

-   **Cohort**: GSE125969, 26 paediatric ependymoma patients, 10x Chromium scRNA-seq (Gillen et al. 2020, *Cell Reports*)
-   **Analytical framing**: re-analysis of the comparator cohort used in Andrade et al. 2024 (*Nat Commun*) to independently characterise the ependymoma TME against the K27M HGG phenotype
-   **Scope of this notebook**:
    -   Read the published count matrix and cell metadata tables
    -   Derive a per-cell `patient_id` from the cell barcode prefix (the published metadata ships without one)
    -   Attach `tumor_subtype` and anatomical `location` as biological covariates (not batch)
    -   Compute QC metrics and apply conservative per-cell filtration
    -   Save a filtered Seurat object as the entry point for downstream integration

# Cohort composition

-   **26 patients across five molecular subgroups**, all linked to tumor subgroup via the shipped `tumor_subtype` column:
    -   **PFA1** (n=12): posterior fossa group A, subtype 1, myeloid-enriched, mesenchymal-prone
    -   **PFA2** (n=7): posterior fossa group A, subtype 2, ciliated-enriched
    -   **PFB** (n=1): posterior fossa group B
    -   **ST-RELA** (n=5): supratentorial RELA-fusion
    -   **ST-YAP** (n=1): supratentorial YAP1-fusion
-   **Location is derivable from subtype prefix** (`PF*` vs `ST*`), no separate column needed
-   **Biological implication for integration**: subtype and location are biological covariates to preserve, not batch effects to remove. `patient_id` is the batch variable Harmony should correct on.

# Checkpoint convention

-   Each notebook loads its predecessor's `.rds` checkpoint and writes its own. Naming follows a step-prefix + short-descriptor scheme so provenance is obvious.

| Stage | File | Description |
|-------------------|------------------|-----------------------------------|
| Raw import (post-metadata merge) | `01-ep-seu.rds` | Seurat object with full cohort, unfiltered |
| Post-QC | `01-ep-seu-filtered.rds` | After cell-level filtration on nFeature, nCount, percent.mt |

# Data import

## Reading the published matrices

-   **Count matrix** (`gse125969_count_matrix.tsv`, \~875 MB): tab-separated, genes as rows, cells as columns. First column holds gene symbols.
-   **Cell metadata** (`gse125969_cell_metadata.tsv`): tab-separated, one row per cell with `cell_id`, `cell_type` (coarse author labels), `tumor_subtype`, and the paper's precomputed UMAP coordinates.
-   **Note**: the paper's UMAP columns are ignored here, I recompute my own embedding after Harmony integration in notebook 02.

```{r}
#| label: read-raw

# Libraries ----------
library(Seurat)
library(tidyverse)
library(readr)
library(Matrix)

# Inputs ----------
counts_raw <- read_tsv(
  "read/gse125969_count_matrix.tsv",
  progress = FALSE,
  show_col_types = FALSE
)
metadata_raw <- read_tsv(
  "read/gse125969_cell_metadata.tsv",
  progress = FALSE,
  show_col_types = FALSE
)

# Processing: coerce counts to a gene x cell sparse matrix ----------
count_matrix <- counts_raw |>
  column_to_rownames(var = colnames(counts_raw)[1]) |>
  as.matrix() |>
  Matrix(sparse = TRUE)

# Sanity ----------
stopifnot(ncol(count_matrix) == nrow(metadata_raw) ||
          length(intersect(colnames(count_matrix), metadata_raw$cell_id)) > 0)
```

## Deriving patient_id from cell barcodes

-   **Problem**: the published metadata has no explicit `patient_id` column. Cell IDs look like `foreman_1239_AAACCTGTCCAGTAGT`, the middle token is the patient's clinical identifier.
-   **Solution**: extract the numeric patient ID with `stringr::str_extract()` using a lookbehind on the `foreman_` prefix.
-   **Cohort verification**: 26 distinct `patient_id` values after extraction, exactly matching the 26 GEO samples (GSM4132013–GSM4132038).
-   **Subtype is already complete**: each patient maps to exactly one `tumor_subtype`, the shipped column is the authoritative link, no external GEO SOFT file lookup required.

```{r}
#| label: derive-patient-id

# Libraries ----------
library(tidyverse)

# Processing ----------
metadata_df <- metadata_raw |>
  mutate(
    patient_id = str_extract(cell_id, "(?<=^foreman_)[0-9]+"),
    location = case_when(
      str_starts(tumor_subtype, "PF") ~ "posterior_fossa",
      str_starts(tumor_subtype, "ST") ~ "supratentorial",
      TRUE ~ NA_character_
    )
  )

# Sanity: 26 patients, each mapped to exactly one subtype ----------
patient_subtype_map <- metadata_df |>
  distinct(patient_id, tumor_subtype, location) |>
  arrange(tumor_subtype, patient_id)

n_patients <- n_distinct(metadata_df$patient_id)
multi_subtype_patients <- patient_subtype_map |>
  count(patient_id) |>
  filter(n > 1)

stopifnot(n_patients == 26L, nrow(multi_subtype_patients) == 0L)

# Output: save the patient table for downstream reference ----------
write_csv(patient_subtype_map, "write/tables/01-patient-subtype-map.csv")
```

| Column | Derivation | Role |
|----------------------|--------------------------------|------------------|
| `patient_id` | `str_extract(cell_id, "(?<=^foreman_)[0-9]+")` | Batch variable for Harmony |
| `tumor_subtype` | As shipped (PFA1 / PFA2 / PFB / ST-RELA / ST-YAP) | Biological covariate, preserve |
| `location` | `PF*` → posterior_fossa, `ST*` → supratentorial | Biological covariate, preserve |
| `cell_type` | As shipped (coarse author labels) | Sanity check only; I re-annotate in 03 |

## Constructing the Seurat object

-   **Cell alignment**: intersect `colnames(count_matrix)` with `metadata_df$cell_id` to drop any cells missing from either table before constructing the Seurat object.
-   **Metadata row order must match cell order** for `CreateSeuratObject()`: set `rownames(metadata_df)` to `cell_id` and reorder.
-   **Project name**: `ep_cellchat`, appears in the object as `seu@project.name`.

```{r}
#| label: build-seurat

# Libraries ----------
library(Seurat)
library(tidyverse)

# Inputs ----------
cells_common <- intersect(colnames(count_matrix), metadata_df$cell_id)

count_matrix_aligned <- count_matrix[, cells_common]
metadata_aligned <- metadata_df |>
  filter(cell_id %in% cells_common) |>
  as.data.frame()
rownames(metadata_aligned) <- metadata_aligned$cell_id
metadata_aligned <- metadata_aligned[colnames(count_matrix_aligned), ]

# Sanity: count columns and metadata rows align after reorder ----------
stopifnot(all(colnames(count_matrix_aligned) == rownames(metadata_aligned)))

# Processing ----------
seu <- CreateSeuratObject(
  counts = count_matrix_aligned,
  meta.data = metadata_aligned,
  project = "ep_cellchat",
  min.cells = 3,
  min.features = 200
)

# Outputs ----------
write_rds(seu, "checkpoints/01-ep-seu.rds")
```

# Quality control

## Computing QC metrics

-   **`percent.mt`**, mitochondrial gene fraction, flags dying / lysing cells leaking cytoplasmic mRNA
-   **`percent.ribo`**, ribosomal protein gene fraction, useful diagnostic for compositional bias but not used for filtration
-   **`log10_genes_per_umi`**, complexity score, lower values indicate under-sequenced or low-information cells

```{r}
#| label: qc-metrics

# Libraries ----------
library(Seurat)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/01-ep-seu.rds")

# Processing ----------
seu[["percent.mt"]]   <- PercentageFeatureSet(seu, pattern = "^MT-")
seu[["percent.ribo"]] <- PercentageFeatureSet(seu, pattern = "^RP[SL]")
seu$log10_genes_per_umi <- log10(seu$nFeature_RNA) / log10(seu$nCount_RNA)

# Outputs ----------
write_rds(seu, "checkpoints/01-ep-seu.rds")
```

## Per-patient QC visualisation

-   **Purpose**: confirm QC distributions are comparable across the 26 patients before applying a cohort-wide threshold
-   **Three violin panels**: `nFeature_RNA`, `nCount_RNA`, `percent.mt`, grouped by `patient_id`, coloured by `tumor_subtype`
-   **What to look for**: patients with systematically lower complexity (e.g. median nFeature \< 500) or elevated mt% should be flagged but not automatically dropped, the brief calls for preserving biological signal

```{r}
#| label: qc-violins
#| fig-width: 16
#| fig-height: 10

# Libraries ----------
library(Seurat)
library(tidyverse)
library(ggplot2)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/01-ep-seu.rds")

# Processing: pull metadata as a plain data frame, then pivot to long form ----------
metadata_df <- seu@meta.data |>
  as_tibble()

qc_long <- metadata_df |>
  select(cell_id, patient_id, tumor_subtype, nFeature_RNA, nCount_RNA, percent.mt) |>
  pivot_longer(
    cols = c(nFeature_RNA, nCount_RNA, percent.mt),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(metric = factor(metric, levels = c("nFeature_RNA", "nCount_RNA", "percent.mt")))

plot_qc_violins <- ggplot(qc_long, aes(x = patient_id, y = value, fill = tumor_subtype)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  labs(x = "Patient", y = "Value", fill = "Subtype") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Outputs ----------
dir.create("write/figures", showWarnings = FALSE, recursive = TRUE)
ggsave(
  filename = "write/figures/01-qc-violins-per-patient.pdf",
  plot = plot_qc_violins,
  width = 16, height = 10, bg = "white"
)
```

## Filtration thresholds

-   **Strategy**: cohort-wide hard floors on `nFeature_RNA`, `nCount_RNA`, and `percent.mt`, simple, auditable, defensible in a ten-minute conversation.

-   **Doublet detection**: deliberately deferred, any doublet cluster will declare itself during annotation in notebook 03.

```{r}
#| label: qc-filter

# Libraries ----------
library(Seurat)
library(tidyverse)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/01-ep-seu.rds")

# Processing: hard-floor QC flags on the metadata (row order == cell order in seu) ----------
metadata_df <- seu@meta.data |>
  as_tibble() |>
  mutate(
    keep = nFeature_RNA >= 200 &
           nCount_RNA   >= 500 &
           percent.mt   <= 20
  )

# Per-patient filtration summary ----------
filter_summary <- metadata_df |>
  group_by(patient_id, tumor_subtype) |>
  summarise(
    cells_before  = n(),
    cells_kept    = sum(keep),
    cells_dropped = cells_before - cells_kept,
    pct_dropped   = round(100 * cells_dropped / cells_before, 1),
    .groups = "drop"
  ) |>
  arrange(tumor_subtype, patient_id)

write_csv(filter_summary, "write/tables/01-qc-filter-summary.csv")

# Apply filter: logical vector aligns 1:1 with seu cells ----------
seu_filtered <- seu[, metadata_df$keep]

# Outputs ----------
write_rds(seu_filtered, "checkpoints/01-ep-seu-filtered.rds")
```

| Metric | Lower cut | Upper cut | Justification |
|------------------|------------------|------------------|--------------------|
| `nFeature_RNA` | ≥200 (hard) + 3-MAD (per-patient) | — | Standard floor for 10x droplet data |
| `nCount_RNA` | ≥500 (hard) + 3-MAD (per-patient) | — | Removes obvious empty droplets and very shallow cells |
| `percent.mt` | — | ≤20 (hard) + 3-MAD (per-patient) | Conservative upper bound for fresh-tissue scRNA-seq |

# Post-filter diagnostics

-   **Per-patient cell yield** is the primary concern before integration, any patient retaining \< 50 cells gets flagged as a candidate for exclusion from CellChat downstream (CellChat communication probability becomes unreliable below \~20 cells per cell type per patient).
-   **Re-plot the QC violins** on the filtered object as the sanity check that the filtration did what the summary table claims.

```{r}
#| label: post-filter-diagnostics
#| fig-width: 14
#| fig-height: 6

# Libraries ----------
library(Seurat)
library(tidyverse)
library(ggplot2)
library(readr)

# Inputs ----------
seu_filtered <- read_rds("checkpoints/01-ep-seu-filtered.rds")

# Processing ----------
metadata_filtered <- seu_filtered@meta.data |>
  as_tibble()

cells_per_patient <- metadata_filtered |>
  dplyr::count(patient_id, tumor_subtype, name = "n_cells") |>
  arrange(tumor_subtype, n_cells)

plot_cells_per_patient <- ggplot(
  cells_per_patient,
  aes(x = reorder(patient_id, n_cells), y = n_cells, fill = tumor_subtype)
) +
  geom_col() +
  coord_flip() +
  labs(x = "Patient", y = "Cells retained", fill = "Subtype") +
  theme_bw(base_size = 11)

# Outputs ----------
ggsave(
  filename = "write/figures/01-cells-per-patient-post-qc.pdf",
  plot = plot_cells_per_patient,
  width = 10, height = 8, bg = "white"
)
```

# Next steps

-   **Notebook 02** loads `01-ep-seu-filtered.rds` and performs `LogNormalize` → variable feature selection (top 2000) → PCA → Harmony on `patient_id` → UMAP → Louvain clustering (k=20), with a pre-vs-post integration UMAP as the integration-success figure required by the brief.
