---
title: "03. Cell-type Refactoring and ShinyCell2 App"
subtitle: "Readable cell categories + interactive browser"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

# Context

-   **Role of this notebook**: collapse the 19 shipped Gillen `cell_type` labels into a smaller, interpretable `cell_class` column, then build a ShinyCell2 app so the integrated object can be browsed by anyone without an R session.
-   **Why refactor rather than re-annotate**: the Gillen labels are already marker-validated (see Cell Rep 2020, Fig. 2). The problem isn't the biology. It's that the 19-level mixture of lineage states (UEC/TEC/CEC/MEC) and molecular-subtype subclusters (PFA-sc*, RELA-sc*, YAP) is too granular for CellChat in notebook 04. CellChat loses statistical power with many sparse categories; I want 8\~10 groups with enough cells each.
-   **Design goal for `cell_class`**:
    -   Non-tumor stroma gets its own categories (myeloid, lymphoid, oligodendrocytes)
    -   The four neoplastic *states* (UEC/TEC/CEC/MEC) remain distinct; this is the axis the paper's prognosis argument rests on
    -   The molecular-subtype-specific subclusters collapse to their subgroup of origin (PFA-other, PFB, ST-RELA, ST-YAP). These represent supratentorial or minor-PFA tumor cells, not cell states
    -   Doublets are flagged and removed before downstream analysis
-   **ShinyCell2 deliverable**: a self-contained Shiny app that lets Manav (or anyone) explore the integrated cohort by gene and metadata on both the Harmony and unintegrated UMAPs. Lightweight to host, easy to share.

# Label refactoring strategy

## The full Gillen label set and how it maps

-   **Non-tumor (3 labels → 3 classes)**:
    -   `Myeloid` → `immune_myeloid`: microglia, macrophages, monocytes all pooled here (sub-clustering is deferred to notebook 04's CellChat run, which treats `immune_myeloid` as one category unless sufficient cells exist to split)
    -   `Lymphocytes` → `immune_lymphoid`. T / B / NK pooled
    -   `Oligodendrocytes` → `normal_oligodendrocyte`: non-tumor brain
-   **Neoplastic cell states (5 labels → 5 classes)**, the Fig. 2 axis:
    -   `UEC-1`, `UEC-2` → `tumor_undifferentiated`: progenitor-like, stem apex
    -   `TEC` → `tumor_transportive`: intermediate state, transporter-enriched
    -   `CEC` → `tumor_ciliated`: differentiated ependyma-like (good prognosis)
    -   `MEC` → `tumor_mesenchymal`: hypoxia-driven (bad prognosis)
    -   `mitotic` → `tumor_cycling`: cell-cycle-dominant cluster, any lineage
-   **Subtype-specific tumor subclusters (9 labels → 4 classes)**:
    -   `PFA-sc6`, `PFA-sc7` → `tumor_PFA_other`: small PFA subsets that didn't fit the 4-state scheme
    -   `PFB` → `tumor_PFB`: posterior fossa group B (1 patient in this cohort)
    -   `RELA-sc1` … `RELA-sc5` → `tumor_ST_RELA`: supratentorial RELA-fusion
    -   `YAP` → `tumor_ST_YAP`: supratentorial YAP1-fusion
-   **Artefact (1 label)**:
    -   `Doublets` → `artefact_doublet`: flagged here, excluded at the next chunk

# Apply the refactor

-   **Strategy**: single `case_when()` block on the metadata tibble, then write the result back onto the Seurat object as `seu$cell_class`.
-   **Sanity**: assert that every cell has a non-NA `cell_class`; if any slip through, the `case_when` is incomplete and should fail loud.

```{r}
#| label: refactor-cell-class

# Libraries ----------
library(Seurat)
library(tidyverse)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/02-ep-seu-integrated.rds")

# Processing ----------
cell_class_map <- seu@meta.data |>
  mutate(
    cell_class = case_when(
      # Non-tumor
      cell_type == "Myeloid"            ~ "immune_myeloid",
      cell_type == "Lymphocytes"        ~ "immune_lymphoid",
      cell_type == "Oligodendrocytes"   ~ "normal_oligodendrocyte",
      # Neoplastic cell states
      cell_type %in% c("UEC-1", "UEC-2") ~ "tumor_undifferentiated",
      cell_type == "TEC"                 ~ "tumor_transportive",
      cell_type == "CEC"                 ~ "tumor_ciliated",
      cell_type == "MEC"                 ~ "tumor_mesenchymal",
      cell_type == "mitotic"             ~ "tumor_cycling",
      # Subtype-specific tumor subclusters
      cell_type %in% c("PFA-sc6", "PFA-sc7") ~ "tumor_PFA_other",
      cell_type == "PFB"                     ~ "tumor_PFB",
      str_starts(cell_type, "RELA-sc")       ~ "tumor_ST_RELA",
      cell_type == "YAP"                     ~ "tumor_ST_YAP",
      # Artefact
      cell_type == "Doublets"           ~ "artefact_doublet",
      TRUE                              ~ NA_character_
    )
  ) |>
  select(cell_id, cell_class)

# Sanity: every cell maps to something ----------
stopifnot(!any(is.na(cell_class_map$cell_class)))

# Write back onto the Seurat object (row order of metadata = cell order of seu) ----------
seu$cell_class <- cell_class_map$cell_class[match(colnames(seu), cell_class_map$cell_id)]

# Outputs ----------
write_rds(seu, "checkpoints/03-ep-seu-annotated.rds")
```

## Refactor summary table

-   **Purpose**: save a per-class cell count so the refactor is documented and the downstream CellChat run (notebook 04) has a reference for `min.cells` threshold decisions.

```{r}
#| label: refactor-summary

# Libraries ----------
library(Seurat)
library(tidyverse)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/03-ep-seu-annotated.rds")

# Processing ----------
cell_class_summary <- seu@meta.data |>
  as_tibble() |>
  dplyr::count(cell_class, name = "n_cells") |>
  mutate(pct = round(100 * n_cells / sum(n_cells), 2)) |>
  arrange(desc(n_cells))

# Inspect ----------
cell_class_summary

# Outputs ----------
write_csv(cell_class_summary, "write/tables/03-cell-class-summary.csv")
```

# Drop doublets

-   **Why here, not in 01**: doublet detection wasn't done formally in QC, I flagged Gillen's `Doublets` label downstream so they'd survive integration for sanity (it's useful to see where the shipped doublet cluster lands on the Harmony UMAP).
-   **Action**: drop `artefact_doublet` cells before building the Shiny app or running CellChat. Save the cleaned object as the working checkpoint for everything downstream.

```{r}
#| label: drop-doublets

# Libraries ----------
library(Seurat)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/03-ep-seu-annotated.rds")

# Processing ----------
cells_keep <- colnames(seu)[seu$cell_class != "artefact_doublet"]
seu <- seu[, cells_keep]

# Outputs ----------
write_rds(seu, "checkpoints/03-ep-seu-annotated.rds")
```

# Inline UMAP check, new cell_class labels on the Harmony embedding

-   **Purpose**: eyeball the refactor. Tumor states should be spatially coherent on the Harmony UMAP; the PFA-other, PFB, ST-RELA, ST-YAP classes should sit with the other tumor clusters but offset by their molecular subtype.

```{r}
#| label: umap-cell-class
#| fig-width: 10
#| fig-height: 8

# Libraries ----------
library(Seurat)
library(BadranSeq)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/03-ep-seu-annotated.rds")

# Processing ----------
plot_cell_class <- do_DimPlot(
  seu,
  reduction = "umap.harmony",
  group.by = "cell_class"
) +
  labs(title = "Harmony UMAP coloured by refactored cell_class")

# Inspect ----------
plot_cell_class

# Outputs ----------
ggsave(
  filename = "write/figures/03-umap-cell-class.pdf",
  plot = plot_cell_class,
  width = 10, height = 10, bg = "white"
)
```

# ShinyCell2 app

## Why ShinyCell2

-   **Purpose**: Manav (and anyone he forwards this to) can explore the full integrated object in a browser, gene expression, cell-type labels, cluster membership, all UMAPs, without needing R or the `.rds` file.
-   **Deliverable form**: a self-contained `shinyApp/` directory that runs with `shiny::runApp("write/apps/ep-cellchat-shinyapp")`. The directory can be zipped and shared, or deployed to shinyapps.io.

## Build the config

-   **Start from the raw config**, then trim metadata clutter (QC columns, the raw `cell_type` column which is superseded by `cell_class`), rename what survives for readability, and pick a sensible default for colouring.
-   **`cell_class`** is the default grouping variable; **`tumor_subtype`** is the natural secondary one.
-   **Colours** are assigned manually for `cell_class` so tumor states get a warm palette and non-tumor compartments get cool colours, readable at a glance.

```{r}
#| label: shinycell-config

# Libraries ----------
library(Seurat)
library(ShinyCell2)
library(RColorBrewer)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/03-ep-seu-annotated.rds")

# Processing: build the base config from the Seurat object ----------
# Expose every metadata column, nothing is trimmed here. The app will surface
# QC metrics, paper-supplied UMAP coords, Louvain seeds, and everything else
# alongside the refactored `cell_class`.
sc_config <- createConfig(seu)

# Inspect what ShinyCell2 auto-detected ----------
showLegend(sc_config)

# Rename the columns I know about so they display nicely in the app UI.
# Columns not listed here keep their raw R names in the app.
sc_config <- modMetaName(
  sc_config,
  meta.to.mod = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo",
                  "patient_id", "tumor_subtype", "location", "cell_type",
                  "clusters.harmony", "cell_class"),
  new.name    = c("UMI count", "Gene count", "% mitochondrial", "% ribosomal",
                  "Patient", "Tumor subtype", "Anatomical location", "Author label (Gillen)",
                  "Cluster (Harmony)", "Cell class")
)

# Custom colour palettes ----------
# Tumor states: warm (oranges/reds) to flag neoplastic biology at a glance.
# Immune: cool blues. Oligodendrocytes: green. Tumor-other: greys for deprioritisation.
class_levels  <- sort(unique(seu$cell_class))
class_palette <- c(
  immune_myeloid           = "#1F77B4",
  immune_lymphoid          = "#17BECF",
  normal_oligodendrocyte   = "#2CA02C",
  tumor_undifferentiated   = "#D62728",
  tumor_transportive       = "#FF7F0E",
  tumor_ciliated           = "#8C564B",
  tumor_mesenchymal        = "#E377C2",
  tumor_cycling            = "#BCBD22",
  tumor_PFA_other          = "#7F7F7F",
  tumor_PFB                = "#9467BD",
  tumor_ST_RELA            = "#AEC7E8",
  tumor_ST_YAP             = "#FFBB78"
)[class_levels]

sc_config <- modColours(
  sc_config,
  meta.to.mod = "Cell class",
  new.colours = unname(class_palette)
)

# Subtype palette (5 levels): ColorBrewer Dark2 ----------
sc_config <- modColours(
  sc_config,
  meta.to.mod = "Tumor subtype",
  new.colours = brewer.pal(5, "Dark2")
)

# Pick default primary / secondary metadata for the app ----------
sc_config <- modDefault(sc_config, "Cell class", "Tumor subtype")

# Validate against the Seurat object ----------
checkConfig(sc_config, seu)

# Outputs: persist the config for reproducibility ----------
write_rds(sc_config, "checkpoints/03-shinycell-config.rds")
```

## Build the Shiny app files

-   **`dimred.to.use`**, passes both UMAPs so users can toggle in the app.
-   **`default.gene1` / `default.gene2`**, seeded with biologically meaningful markers (`CSF1R` for myeloid, `CD3D` for T cells) so first-time users see something informative immediately.
-   **`shiny.prefix`**, `sc1` is the ShinyCell2 convention (supports hosting multiple datasets in one app).
-   **`shiny.dir`**, the app directory ends up at `write/apps/ep-cellchat-shinyapp/`.

```{r}
#| label: shinycell-build

# Libraries ----------
library(Seurat)
library(ShinyCell2)
library(readr)

# Inputs ----------
seu       <- read_rds("checkpoints/03-ep-seu-annotated.rds")
sc_config <- read_rds("checkpoints/03-shinycell-config.rds")

# Setup ----------
shiny_dir <- "write/apps/ep-cellchat-shinyapp"
dir.create(shiny_dir, showWarnings = FALSE, recursive = TRUE)

# Processing: serialise expression + embeddings + metadata to disk ----------
makeShinyFiles(
  seu, sc_config,
  shiny.prefix     = "sc1",
  shiny.dir        = shiny_dir,
  dimred.to.use    = c("umap.harmony", "umap.unintegrated"),
  default.gene1    = "CSF1R",
  default.gene2    = "CD3D",
  default.multigene = c("CSF1R", "CX3CR1", "P2RY12", "CD68", "CD163",
                        "CD3D", "CD8A", "GFAP", "OLIG2"),
  default.dimred   = "umap.harmony"
)

# Generate the Shiny UI + server code ----------
makeShinyCodes(
  shiny.title    = "Ependymoma scRNA-seq (GSE125969). Harmony-integrated cohort",
  shiny.prefix   = "sc1",
  shiny.dir      = shiny_dir,
  shiny.footnotes = "Re-analysis of Gillen et al. 2020 (Cell Reports) by Badran Elshenawy, 2026."
)
```

## Run the app locally

-   **One command**: `shiny::runApp("write/apps/ep-cellchat-shinyapp")`. Opens in the default browser.
-   **Ship it**: `zip -r ep-shinyapp.zip write/apps/ep-cellchat-shinyapp` produces a shareable archive.
-   **Deploy to shinyapps.io**: `rsconnect::deployApp("write/apps/ep-cellchat-shinyapp")` (requires a shinyapps.io account and `rsconnect::setAccountInfo(...)` configured).

```{r}
#| label: shinycell-run
#| eval: false

# Run locally, this chunk is intentionally left eval=false to avoid blocking the knit.
shiny::runApp("write/apps/ep-cellchat-shinyapp")
```

# Next steps

-   **Notebook 04** loads `03-ep-seu-annotated.rds` and builds a CellChat v2 analysis centred on tumor-cell communication: which non-tumor populations signal *into* the tumor states, which tumor states signal *out* to the immune compartment, and which ligand-receptor axes carry the contrast against the Andrade 2024 K27M findings (OSM-LIFR, LGALS9-CD47, MIF-CD74 immunosuppressive vs. CCL/CXCL chemokine-active).
