---
title: "05. Hallmark Module Scoring and Shiny App Rebuild"
subtitle: "Seurat AddModuleScore + MSigDB Hallmark, then ShinyCell2 v2"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

# Context

-   **Role of this notebook**: the Pillar 3 deliverable. Hallmark-programme scoring of every cell in the integrated cohort, followed by a rebuild of the ShinyCell2 app so the 50 scores are browsable as metadata colour variables.
-   **Method choice, `Seurat::AddModuleScore` instead of `escape::runEscape`**: the `escape` + `msigdb` stack has been fragile on this machine, and for 50 well-curated gene sets on \~16k cells, Seurat's module-score approach gives nearly identical ranking of cell states while being faster, simpler, and dependency-free. Scores are stored as per-cell metadata columns rather than an assay, which also simplifies ShinyCell integration.
-   **`msigdbr` for gene sets**: tidy CRAN data frame of the full MSigDB, trivial to split into the named-list format `AddModuleScore` expects (`features = list(pathway_1 = c(gene symbols), ...)`).
-   **Upstream**: loads `03-ep-seu-annotated.rds`. Independent of notebook 04 (CellChat).

# Checkpoint convention

| Stage | File | Description |
|------------------------|------------------------|------------------------|
| Post-scoring | `checkpoints/05-ep-seu-hallmark.rds` | Seurat object with 50 Hallmark score columns in metadata |
| Shiny config | `checkpoints/05-shinycell-config.rds` | Rebuilt ShinyCell2 config |
| App directory | `write/apps/ep-cellchat-shinyapp-v2/` | Standalone rebuilt Shiny app |

# Fetch Hallmark gene sets from MSigDB

-   **`msigdbr(db_species = "HS", collection = "H")`** pulls the 50 Hallmark pathways. Returns a long-format tibble with `gs_name` and `gene_symbol` columns.
-   **Refactor to named list**: `split(df$gene_symbol, df$gs_name) |> lapply(unique)`, ready for `AddModuleScore(features = …)`.

```{r}
#| label: gene-sets

# Libraries ----------
library(msigdbr)
library(dplyr)
library(readr)

# Processing ----------
hallmark_df <- msigdbr(db_species = "HS", collection = "H")

hallmark_sets <- split(hallmark_df$gene_symbol, hallmark_df$gs_name)
hallmark_sets <- lapply(hallmark_sets, unique)

cat("Hallmark pathways loaded: ", length(hallmark_sets), "\n")
cat("Average genes per pathway:", round(mean(lengths(hallmark_sets)), 1), "\n")
cat("First 5 names:\n")
print(names(hallmark_sets)[1:5])

# Outputs ----------
write_rds(hallmark_sets, "checkpoints/05-hallmark-genesets.rds")
```

# Run AddModuleScore across all cells

-   **Why it works for this**: `AddModuleScore` computes, for each cell, the average expression of the gene set minus the average expression of 100 matched control genes binned by expression level. Robust, interpretable, and aligned to Seurat's ecosystem.
-   **Parameters**:
    -   `features = hallmark_sets`, the named list from msigdbr
    -   `name = "Hallmark_"`. Seurat suffixes each set with its index (`Hallmark_1`, …, `Hallmark_50`), which I rename to the actual pathway name immediately after
    -   `ctrl = 100`, number of control genes per bin (default)
    -   `nbin = 24`, expression bins (default)
    -   `seed = 1`, reproducibility
-   **Post-processing**: rename the `Hallmark_*` metadata columns to the real pathway IDs so downstream plots and the Shiny app show meaningful labels.
-   **Runtime**: \~1–2 min for 50 sets on \~16k cells.

```{r}
#| label: module-scores

# Libraries ----------
library(Seurat)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/03-ep-seu-annotated.rds")
hallmark_sets <- read_rds("checkpoints/05-hallmark-genesets.rds")

DefaultAssay(seu) <- "RNA"

# Processing ----------
seu <- AddModuleScore(
  seu,
  features = hallmark_sets,
  name     = "Hallmark_",
  ctrl     = 100,
  nbin     = 24,
  seed     = 1
)

# Rename Hallmark_1, Hallmark_2, ... -> actual pathway names.
old_cols <- paste0("Hallmark_", seq_along(hallmark_sets))
new_cols <- names(hallmark_sets)

stopifnot(all(old_cols %in% colnames(seu@meta.data)))
colnames(seu@meta.data)[match(old_cols, colnames(seu@meta.data))] <- new_cols

cat("Hallmark score columns added:", length(new_cols), "\n")
cat("First 3 columns:\n")
print(head(seu@meta.data[, new_cols[1:3]]))

# Outputs ----------
write_rds(seu, "checkpoints/05-ep-seu-hallmark.rds")
```

# Cohort-wide Hallmark heatmap

-   **Purpose**: one figure summarising which pathways are active in which cell class.
-   **Build**: mean score per pathway × cell_class, z-scaled row-wise to make columns (cell classes) directly comparable per pathway.
-   **Library**: `ComplexHeatmap` for annotation and dendrogram support.

```{r}
#| label: hallmark-heatmap-cellclass
#| fig-width: 16
#| fig-height: 14

# Libraries ----------
library(Seurat)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/05-ep-seu-hallmark.rds")

hallmark_cols <- grep("^HALLMARK_", colnames(seu@meta.data), value = TRUE)

# Processing ----------
score_long <- seu@meta.data |>
  select(cell_id, cell_class, all_of(hallmark_cols)) |>
  pivot_longer(cols = all_of(hallmark_cols),
               names_to = "pathway", values_to = "score")

score_matrix <- score_long |>
  group_by(pathway, cell_class) |>
  summarise(mean_score = mean(score), .groups = "drop") |>
  pivot_wider(names_from = cell_class, values_from = mean_score) |>
  column_to_rownames("pathway") |>
  as.matrix()

# Z-score each pathway across cell classes for comparable colour scale.
score_matrix_z <- t(apply(score_matrix, 1, function(x) (x - mean(x)) / sd(x)))

col_fun <- colorRamp2(c(-2, 0, 2), c("#2166AC", "#F7F7F7", "#B2182B"))

plot_heatmap <- Heatmap(
  score_matrix_z,
  name          = "Z-score",
  col           = col_fun,
  row_names_gp  = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 10),
  column_names_rot = 45,
  cluster_columns = TRUE,
  cluster_rows    = TRUE,
  row_title     = "Hallmark pathway",
  column_title  = "Cell class"
)

# Inspect ----------
draw(plot_heatmap)

# Outputs ----------
dir.create("write/figures", showWarnings = FALSE, recursive = TRUE)
pdf("write/figures/05-hallmark-heatmap-cellclass.pdf", width = 16, height = 14)
draw(plot_heatmap)
dev.off()
png("write/figures/05-hallmark-heatmap-cellclass.png",
    width = 16 * 120, height = 14 * 120, res = 150)
draw(plot_heatmap)
dev.off()

write_csv(
  score_matrix |> as.data.frame() |> rownames_to_column("pathway"),
  "write/tables/05-hallmark-mean-by-cell-class.csv"
)
```

# Focused pathway panels

-   **Panel 1. Myeloid inflammation + metabolism**: `Seurat::VlnPlot` across `cell_class` for the 5 inflammation Hallmarks and 3 metabolism Hallmarks. Visualises whether ependymoma myeloid cells are inflammation-dominant (the Andrade-expected pattern).
-   **Panel 2. Tumor state × hypoxia**: subset to PFA tumor states (UEC/TEC/CEC/MEC), ridge plot of `HALLMARK_HYPOXIA`. MEC should dominate.
-   **Panel 3. Proliferation axis**: G2M / E2F / mitotic spindle Hallmarks, should peak in undifferentiated and cycling tumor classes.
-   **Plot approach**: `Seurat::VlnPlot` and `Seurat::RidgePlot` both accept metadata column names as `features`, so the scores plot directly without any assay gymnastics.

```{r}
#| label: inflammation-panel
#| fig-width: 16
#| fig-height: 10

# Libraries ----------
library(Seurat)
library(patchwork)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/05-ep-seu-hallmark.rds")

# Processing ----------
inflam_metab <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_COMPLEMENT",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_HYPOXIA",
  "HALLMARK_GLYCOLYSIS",
  "HALLMARK_OXIDATIVE_PHOSPHORYLATION"
)

plot_inflam_metab <- VlnPlot(
  seu,
  features = inflam_metab,
  group.by = "cell_class",
  pt.size  = 0,
  ncol     = 2,
  same.y.lims = FALSE
) &
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Inspect ----------
plot_inflam_metab

# Outputs ----------
ggsave(
  filename = "write/figures/05-hallmark-inflammation-metabolism-panel.pdf",
  plot = plot_inflam_metab,
  width = 16, height = 12, bg = "white",
  device = cairo_pdf
)
ggsave(
  filename = "write/figures/05-hallmark-inflammation-metabolism-panel.png",
  plot = plot_inflam_metab,
  width = 16, height = 12, bg = "white", dpi = 150
)
```

```{r}
#| label: tumor-state-hypoxia
#| fig-width: 10
#| fig-height: 6

# Libraries ----------
library(Seurat)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/05-ep-seu-hallmark.rds")

# Processing ----------
tumor_states <- c("tumor_undifferentiated", "tumor_transportive",
                  "tumor_ciliated",         "tumor_mesenchymal")
seu_tumor <- seu[, seu$cell_class %in% tumor_states]
seu_tumor$cell_class <- factor(seu_tumor$cell_class, levels = tumor_states)

plot_hypoxia_ridge <- RidgePlot(
  seu_tumor,
  features = "HALLMARK_HYPOXIA",
  group.by = "cell_class"
) +
  labs(title = "Hypoxia score across PFA tumor states (MEC expected to dominate)")

# Inspect ----------
plot_hypoxia_ridge

# Outputs ----------
ggsave(
  filename = "write/figures/05-hallmark-hypoxia-tumor-states.pdf",
  plot = plot_hypoxia_ridge,
  width = 10, height = 6, bg = "white",
  device = cairo_pdf
)
ggsave(
  filename = "write/figures/05-hallmark-hypoxia-tumor-states.png",
  plot = plot_hypoxia_ridge,
  width = 10, height = 6, bg = "white", dpi = 150
)
```

```{r}
#| label: proliferation-panel
#| fig-width: 12
#| fig-height: 10

# Libraries ----------
library(Seurat)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/05-ep-seu-hallmark.rds")

# Processing ----------
prolif_pathways <- c(
  "HALLMARK_G2M_CHECKPOINT",
  "HALLMARK_E2F_TARGETS",
  "HALLMARK_MITOTIC_SPINDLE"
)

plot_prolif <- VlnPlot(
  seu,
  features = prolif_pathways,
  group.by = "cell_class",
  pt.size  = 0,
  ncol     = 1
) &
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Inspect ----------
plot_prolif

# Outputs ----------
ggsave(
  filename = "write/figures/05-hallmark-proliferation-panel.pdf",
  plot = plot_prolif,
  width = 12, height = 10, bg = "white",
  device = cairo_pdf
)
ggsave(
  filename = "write/figures/05-hallmark-proliferation-panel.png",
  plot = plot_prolif,
  width = 12, height = 10, bg = "white", dpi = 150
)
```

# Feature-on-UMAP. Hypoxia score, Harmony embedding

-   **Purpose**: one publication-friendly UMAP coloured by the MEC-defining pathway, so MEC's spatial coherence with hypoxia is visible at a glance.

```{r}
#| label: hypoxia-featureplot
#| fig-width: 8
#| fig-height: 7

# Libraries ----------
library(Seurat)
library(BadranSeq)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/05-ep-seu-hallmark.rds")

# Processing ----------
plot_hypoxia_umap <- do_FeaturePlot(
  seu,
  reduction = "umap.harmony",
  features  = "HALLMARK_HYPOXIA",
  pt.size   = 0.4,
  plot_cell_borders = FALSE
) +
  labs(title = "Hallmark hypoxia score on Harmony UMAP")

# Inspect ----------
plot_hypoxia_umap

# Outputs ----------
ggsave(
  filename = "write/figures/05-umap-harmony-hypoxia.pdf",
  plot = plot_hypoxia_umap,
  width = 8, height = 7, bg = "white",
  device = cairo_pdf
)
ggsave(
  filename = "write/figures/05-umap-harmony-hypoxia.png",
  plot = plot_hypoxia_umap,
  width = 8, height = 7, bg = "white", dpi = 150
)
```

# ShinyCell2, rebuild the app with Hallmark scores

-   **Key point**: the 50 Hallmark scores live as continuous metadata columns on the Seurat object, `createConfig(seu)` picks them up automatically. They become queryable as colour variables in the metadata panel of the app.
-   **App directory**: `write/apps/ep-cellchat-shinyapp-v2/`, parallel to the v1 from notebook 03, so both survive.
-   **Default**: seed the landing view with `cell_class` × `tumor_subtype` as before; users can swap in any Hallmark pathway from the metadata dropdown.

```{r}
#| label: shinycell-rebuild

# Libraries ----------
library(Seurat)
library(ShinyCell2)
library(RColorBrewer)
library(readr)

# Inputs ----------
seu <- read_rds("checkpoints/05-ep-seu-hallmark.rds")

# Setup ----------
shiny_dir <- "write/apps/ep-cellchat-shinyapp-v2"
dir.create(shiny_dir, showWarnings = FALSE, recursive = TRUE)

# Processing ----------
sc_config <- createConfig(seu)

# Rename the columns I have friendly labels for.
sc_config <- modMetaName(
  sc_config,
  meta.to.mod = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo",
                  "patient_id", "tumor_subtype", "location", "cell_type",
                  "clusters.harmony", "cell_class"),
  new.name    = c("UMI count", "Gene count", "% mitochondrial", "% ribosomal",
                  "Patient", "Tumor subtype", "Anatomical location", "Author label (Gillen)",
                  "Cluster (Harmony)", "Cell class")
)

# Custom palette for Cell class ----------
class_levels  <- sort(unique(seu$cell_class))
class_palette <- c(
  immune_myeloid          = "#1F77B4",
  immune_lymphoid         = "#17BECF",
  normal_oligodendrocyte  = "#2CA02C",
  tumor_undifferentiated  = "#D62728",
  tumor_transportive      = "#FF7F0E",
  tumor_ciliated          = "#8C564B",
  tumor_mesenchymal       = "#E377C2",
  tumor_cycling           = "#BCBD22",
  tumor_PFA_other         = "#7F7F7F",
  tumor_PFB               = "#9467BD",
  tumor_ST_RELA           = "#AEC7E8",
  tumor_ST_YAP            = "#FFBB78"
)[class_levels]

sc_config <- modColours(sc_config,
                        meta.to.mod = "Cell class",
                        new.colours = unname(class_palette))
sc_config <- modColours(sc_config,
                        meta.to.mod = "Tumor subtype",
                        new.colours = brewer.pal(5, "Dark2"))

sc_config <- modDefault(sc_config, "Cell class", "Tumor subtype")
checkConfig(sc_config, seu)

write_rds(sc_config, "checkpoints/05-shinycell-config.rds")

# Build shiny files ----------
makeShinyFiles(
  seu, sc_config,
  shiny.prefix      = "sc1",
  shiny.dir         = shiny_dir,
  dimred.to.use     = c("umap.harmony", "umap.unintegrated"),
  default.gene1     = "CSF1R",
  default.gene2     = "CD3D",
  default.multigene = c("CSF1R", "CX3CR1", "P2RY12", "CD68", "CD163",
                        "CD3D", "CD8A", "GFAP", "OLIG2"),
  default.dimred    = "umap.harmony"
)

makeShinyCodes(
  shiny.title     = "Ependymoma scRNA-seq (GSE125969). Harmony + Hallmark module scores",
  shiny.prefix    = "sc1",
  shiny.dir       = shiny_dir,
  shiny.footnotes = "Re-analysis of Gillen et al. 2020 (Cell Reports) with Hallmark AddModuleScore. Badran Elshenawy, 2026."
)
```

## Run the rebuilt app locally

```{r}
#| label: shinycell-run-v2
#| eval: false

shiny::runApp("write/apps/ep-cellchat-shinyapp-v2")
```

# Next steps

-   **Notebook 06** (summary / writeup) pulls concrete numbers from 04 (CellChat tables) and 05 (Hallmark mean scores) into the 500–1000 word biological report the brief specifies.
